---
title: "Calculating vertical distance from drainage"
author: "Lorena Abad"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: 
  md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE, eval = F)
```

# Calculating vertical distance from drainage

For water mapping purposes, topographic properties of the study area are well received to remove false positives. These properties are derived from Digital Elevation Models (DEM). One important derived property is the vertical distance from drainage, which, depending on the software, can be computed with different algorithms. One of them is the Height above the Nearest Drainage (HAND), computed in TauDEM or Hydroshare. Another option is the Vertical Distance to Channel Network computed in SAGA. Since the `RSAGA` package allows an easy bridge between R and SAGA, I will focus on this last algorithm. 

## Stepwise approach:

### 1. Load data into R
Since SAGA works with data on disk, we will load the original data as a raster object, and then save it as a SAGA `.sgrd` file. 

```{r, eval = T}
library(raster)
dem <- raster('../../NZ_DEM_15m/nzdem_nztm_nzgd49_kaikoura.tif') 
```

```{r}
writeRaster(dem, filename = 'dem.sgrd', format = 'SAGA', overwrite = T, NAflag = 0)
```


### 2. Call SAGA environment
```{r}
library(RSAGA)
env <- rsaga.env()
```

### 3. Fill voids on DEM
```{r}
rsaga.fill.sinks(in.dem = "dem.sgrd", out.dem = "dem_filled", method = "wang.liu.2006", env = env)
```

Plot the data with stars:
```{r, eval = T}
library(stars)
dem_filled <- read_stars('dem_filled.sdat', proxy = T)
plot(dem_filled)
```

### 4. Compute flow accumulation
```{r}
rsaga.get.modules('ta_hydrology', env = env)
rsaga.get.usage('ta_hydrology', 0, env = env)
rsaga.geoprocessor(
  'ta_hydrology', 0, 
  list(ELEVATION='dem_filled.sgrd', FLOW='flow_accummulation', METHOD=0),
  env = env
)
```

```{r, eval = T}
library(stars)
flow_accummulation <- read_stars('flow_accummulation.sdat', proxy = T)
plot(flow_accummulation)
```

### 5. Calculate channel network
```{r}
rsaga.get.modules('ta_channels', env = env)
rsaga.get.usage('ta_channels', 0, env = env)
rsaga.geoprocessor(
  'ta_channels', 0, 
  list(ELEVATION='dem_filled.sgrd', INIT_GRID='flow_accummulation.sgrd', INIT_VALUE=1000000, CHNLNTWRK="channel_net.sdat",CHNLROUTE="route.sdat",SHAPES='channel_net_shp.shp'),
  env = env
)
```

```{r, eval = T}
library(sf)
channel_net <- st_read('channel_net_shp.shp')
plot(channel_net['Order'])
```

### 6. Calculate vertical distance to channel network
```{r}
rsaga.get.usage('ta_channels', 3, env = env)
rsaga.geoprocessor(
  'ta_channels', 3, 
  list(ELEVATION='dem_filled.sgrd', CHANNELS='channel_net.sgrd', DISTANCE='vdcn'),
  env = env
)
```

```{r, eval = T}
vdcn <- read_stars('vdcn.sdat')
vdcn <- st_set_crs(vdcn, value = st_crs(dem))
plot(vdcn)
```

```{r}
write_stars(vdcn, 'vdcn.tiff')
```

